### 2012-01-23 paper_draft.tex
New structure for the sharing genotypes and the crowdsourcing phenotypes section. The verbatim package is used to create multi-line comments

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

@@

The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers. DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers or any other reuse in a scientific context.

\subsubsection*{Sharing genotypic information}
Currently users can upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints.
Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed up with openSNP, 79 of which have uploaded their genotyping files. The openSNP database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.
\subsubsection*{Crowdsourcing phenotypes}
Users are encouraged to list as many phenotypes as possible through a simple achievement system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile pages. The specification of variations on a single phenotype and the phenotypes themselves is open and not limited to pre-defined categories. Every user can add completely new phenotypes that are not yet queried by openSNP. To reduce the amount of manual data curation, openSNP tries to harmonize the expression and spelling of the same phenotype or variation by implementing an autocompletion-feature which lists similar entries to what is being entered that are already in the openSNP database. In the same timeframe all users combined have entered a total of 675 variations on 47 different phenotypes. See figure n for a distribution of data acquisition over time.
The mean number of users that have entered their variations for a single phenotype is 14.36 (SD 12.65), the median is 10. The distribution of how many users have entered their data per phenotype can be seen in figure n+1. The phenotype provided by the most users is the eye color, which has been entered by 54 users. There are two phenotypes which have so far only been provided by a single user: The SAT Writing score and triglyceride-levels.
\subsubsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, SNPs are ranked by the amount of information gathered by the external services.The external services themselves are ranked by how easily users can access information from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.  
\subsubsection*{Data access}
OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. Additionally, the genetic data is accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. Users can discuss SNPs and phenotypes on the platform using a simple commenting system or private messages.

>>>>

%FZ: Enable the comment command
\usepackage{verbatim}
% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

@@

The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers. DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers or any other reuse in a scientific context. 
\begin{comment}
FZ: That is a good summary, but as Bastian pointed out the reuse is not only
enabled in a scientific context?
\end{comment}

\begin{comment}
FZ:
\end{comment}
\subsubsection*{Sharing genotypic information}
Currently users can upload their genotypings from the companies 23andme, deCODEme and Family Tree to/via a webinterface to the openSNP 
project. The uploaded data is published under the Creative Commons Zero-license, 
which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - 
allows a complete reuse of the data without any constraints.
\begin{comment}
FZ: If that is true, rephrase sentence above
FZ: Made two sentences out of one.
\end{comment}
Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed 
up with openSNP, 79 of which have uploaded their genotyping files. The openSNP 
database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs.

\subsubsection*{Crowdsourcing phenotypes}
Beside uploading genotypic data, users may list new phenotypes that are not yet 
queried by openSNP. The specification of these phenotypes  is open and not limited 
to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
the expression and spelling of the same phenotype or variation. We implemented an 
autocompletion-feature which helps users to reuse already entered phenotypes.
Users are encouraged to list as many phenotypes as possible through a simple 
achievement system which rewards users that upload their data and enter phenotypic 
information with small badges that are shown on their profile pages.
\begin{comment}
FZ: I changed the structure of the paragraph to make it a little bit more
concise.
I think its best to connect the phenotypes to the above paragraph, explain
the procedure and then talk about the achievments
\end{comment}
In the same timeframe that was mentioned above, all users combined have 
entered a total of 675 variations on 47 different phenotypes. See figure n for a 
distribution of data acquisition over time.
The mean number of users that have entered their variations for a single phenotype 
is 14.36 (SD 12.65), the median is 10. The distribution of how many users have 
entered their data per phenotype can be seen in figure n+1. The phenotype provided 
y the most users is the eye color, which has been entered by 54 users. There are 
two phenotypes which have so far only been provided by a single user: 
The SAT Writing score and triglyceride-levels.
\begin{comment}
FZ: I think we should clarify the difference between variaton and phenotype!
\end{comment}

\subsubsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, SNPs are ranked by the amount of information gathered by the external services.The external services themselves are ranked by how easily users can access information from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.  
\subsubsection*{Data access}
OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. Additionally, the genetic data is accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. Users can discuss SNPs and phenotypes on the platform using a simple commenting system or private messages.
